% ---
% Agradecimentos
% ---


First of all, I would like to thank all my friends, from my undergraduate degree at Unicamp, from my double diploma at Telecom ParisTech, from my year of doctorate sandwich at Cedars-Sinai, from the handball teams. I ask my sincere excuses if I have failed them during these last years, but I will eternally thank for all the support, for all the laughter, for all the hugs, for all understanding in those my depressive days.

I would like to thank my parents for all their support in these moments of indecision, for being more understanding with each passing day, for always supporting me to pursue my dreams and for all the love they have always given me. Thanks even for the moments of scolding, because I know they were all to make me a better person.

I also would like to thank my sister for teaching me to never doubt yourself,
for all conversations, for all trips together, to be a great friend and a wonderful sister.

Notably, I would like to thank my great friend, Dr. Hona, for supporting my family these last years.
I still remember the first day I met her. She was in a wheelchair, her health was not good, despite all this she had that incredible smile on her face. With whom I have had the most  profound conversations, she has always given me her extremely knowledgeable advice. I do not consider only a great friend, but an enormous inspiration as a person. I thank her for reminding me that the noblest thing a man can do in his life is to help others; that happiness lies not in the great moments of life, but in those little quotidian moments that we forget to enjoy. In these darkest moments of humanity, people like her brought me hope of a better world. Her recent passing has brought us sad times, but she will forever remain a part of our lives. Thank you so much for everything.

I would like to thank  Dr. Bernardo Mantovani for being a great inspiration and for teaching me the importance of the Public University in society. I'm also grateful to my scientific initiation mentor, Dra. Islene Calciolari, for all the support when everything went wrong, and for being an excellent teacher and friend.

Finally, I also would like to thank FAPESP for the financial support,
which was fundamental to carry out all the research activities.
And to my advisers Dr. Houtan Noushmehr and Dr. Benjamin Berman
for all guidance and support during this project.\\
